---
title: "Analysis of Causes of Homeless Deaths"
subtitle: "My subtitle if needed"
author: 
  - Juliana Zhu
thanks: "Code and data are available at: LINK."
date: 26th September 2024
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
```


# Introduction

The homeless are one of the most vulnerable groups in society, and there are many factors that seriously affect their health and well-being. According to the data on the causes of death among the homeless from OpenDataToronto, the main causes of death among the homeless from 2017 to 2023 are Accident, Cancer, Cardiovascular Disease, COVID-19, Drug toxicity, Homicide, Infection, Pneumonia, and Suicide. By analyzing the data on the causes of death, we can better understand which factors are most harming the health and well-being of the homeless so that we can call on governments and society to take measures.We use @citeR and @rohan.

The remainder of this paper is structured as follows. @sec-data....



# Data {#sec-data}

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it. 






\newpage


# References


